### 2023-08-29 proofs/main.tex
WIP: update typing lemmas

The signature $[[Θ ⊢ uσ]]$ means that $[[uσ]]$ maps each algorithmic 
variable appearing in $[[Θ]]$ to a type well-formed in the corresponding
context; and for each variable not appearing in $[[dom(Θ)]]$, acts as identity.

\begin{definition}[Signature of Algorithmic Substitution]
  $[[Θ ⊢ uσ]]$ means that
  \begin{enumerate}
    \item for any $[[ α̂±[Γ] ∊ Θ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$ and
    \item for any $[[ α̂± ∉ dom(Θ)]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
  \end{enumerate}
\end{definition}

Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.
In contrast to algorithmic substitution $[[uσ]]$,
it only defined on the negative algorithmic variables, 
and it allows mapping algorithmic variables to
\emph{algorithmic} terms.

>>>>

The signature $[[Θ ⊢ uσ : Ξ]]$ means that
$[[Ξ ⊆ dom(Θ)]]$ and 
$[[uσ]]$ maps each algorithmic variable 
from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
and for each variable not appearing in $[[dom(Θ)]]$, 
it acts as identity.

\begin{definition}[Signature of Algorithmic Substitution]
  $[[Θ ⊢ uσ : Θ]]$ means that
  \begin{enumerate}
    \item for any $[[α̂± ∊ Ξ]]$,
      there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
      and $[[ Γ ⊢ [uσ]α̂± ]]$; 
    \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
  \end{enumerate}
\end{definition}

Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.
In contrast to algorithmic substitution $[[uσ]]$,
it allows mapping algorithmic variables to
\emph{algorithmic} terms.

### 2023-08-29 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: update typing lemmas

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
        $[[Θ ⊢ SC]]$ and 
        for any normalized $[[uσ]]$ such that $[[(Θ ⊢ SC) ⊢ uσ]]$,
        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.

    Suppose that $[[uσ]]$ is normalized and $[[(Θ ⊢ SC) ⊢ uσ]]$,
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item \ruleref{\ottdruleAArrowLabel}. Then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ SC]]$\\
        On the next step, the the algorithm makes two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ SC2]]$
        and returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.

        By the soundness of constraint merge (\cref{lemma:merge-soundness}),
        $[[(Θ ⊢ SC1) ⊢ uσ]]$ and $[[(Θ ⊢ SC2) ⊢ uσ]]$.
        Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
        and by the induction hypothesis, $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
        This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [uσ](uP → uN') ≤ iQ → iM']]$.

@@

\begin{lemma}[Completeness of the Negative Subtyping]
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [uσ]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

@@

        This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$.
        Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[(Θ  ⊢  SC1) ⊢ uσ]]$.

        Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
        satisfies the required properties.
        To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[(Θ  ⊢  SC) ⊢ uσ]]$, as required.

>>>>

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
        $[[Θ ⊢ SC : uv(uN)]]$ and 
        for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.

    Suppose that $[[uσ]]$ is normalized and $[[ Θ  ⊢ uσ : SC ]]$,
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item \ruleref{\ottdruleAArrowLabel}. Then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ SC]]$\\
        On the next step, the the algorithm makes two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ SC2]]$
        and returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.

        By the soundness of constraint merge (\cref{lemma:merge-soundness}),
        $[[ Θ  ⊢ uσ : SC1 ]]$ and $[[ Θ  ⊢ uσ : SC2 ]]$.
        Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
        and by the induction hypothesis, $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
        This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [uσ](uP → uN') ≤ iQ → iM']]$.

@@

\begin{lemma}[Completeness of the Negative Subtyping]
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
            Then $[[ uN ]] = [[ ↑uP ]]$, since
            the substitution $[[ [uσ]uN ]]$ must preserve the 
            top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
            and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
            The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
            \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
            and the difference in the induction statement for the positive and 
            the negative case here does not matter.

@@

            This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv(uN')]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
            Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv(uP)]]$
            and $[[ Θ ⊢ uσ : SC1 ]]$.

            Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
            satisfies the required properties.
            To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness})
            (notice that $[[Θ ⊢ uσ : uv(uP → uN')]]$ means 
            $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$).
            This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined and $[[ Θ ⊢ uσ : SC ]]$ holds.
